---
title: "P1_Data_Clean"
author: "mtzivano"
date: "2023-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Clean Arthropod Data from GBIF and other sources

For each of the taxonomic groups included in the project, data are assembled and cleaned to enter the modeling pipeline. This cleaning procedure incudes the following steps:

* Aggregate data from different sources
* Resolve potential taxonomic issues
* Ensure that datasets are not overlapping
* Remove suspect records using CoordinateCleaner
* Rasterize on grid
* Report on summary statistics for the group, including number of species, number of records initially and after each cleaning step, maps of sampling effort/species richness and environmental coverage of observations

### Setup

```{r include=FALSE}
# Libraries
library(stringr)
library(data.table)
library(raster)
library(terra)
library(tidyverse)
library(countrycode)
library(CoordinateCleaner)
library(ggplot2)
library(progress)
```

```{r}
# Source functions
source("C:/Users/tzivanom/Documents/Scripts/Functions/F1_CleanData_FUNCTIONS.R")

# Paths
path_species_data <- "D:/GBIF_INVERTEBRATE_RAW/GBIF_extracted_data/"
path_GIS_data <- "C:/Users/tzivanom/Documents/Data/GIS/"
path_save <- "D:/PostFiltering/GBIF_only/"
path_save_obs <- "D:/PostFiltering/OBSorg_only/"
path_final <- "D:/PostFiltering/Final_Rasters_AllData/"
path_results <- "C:/Users/tzivanom/Documents/Results/NaturaConnect/DataCleaning/"
```

```{r}
# References

# NaturaConnect raster template
NaturaConnect_grid1k <- raster(paste0(path_GIS_data, "BIOCLIMP_NC/EUR11_1991-2020.tif"))[[1]]
NaturaConnect_grid_r <- rast(NaturaConnect_grid1k)
values(NaturaConnect_grid_r) <- NA

# EUROSTAT country references
EUROSTAT_Countries_1k <- shapefile(paste0(path_GIS_data, "CountryBorders_EUROSTAT/EUROSTAT_Countries_buffered1k_WGS84.shp"))
```


```{r}
# Group variables
taxo_groups_all <- c("Insecta/Coleoptera/Carabidae",
                     "Insecta/Hymenoptera/Formicidae",
                     "Arachnida/Araneae",
                     "Gastropoda",
                     "Insecta/Odonata",
                     "Insecta/Hymenoptera/Ichneumonoidea",
                     "Insecta/Lepidoptera/Heterocera",
                     "Insecta/Orthoptera",
                     "Collembola",
                     "Clitellata/Crassiclitellata",
                     "Insecta/Lepidoptera/Rhopalocera",
                     "Insecta/Diptera/Syrphidae",
                     "Insecta/Hymenoptera/Anthophila")

taxo_groups_gbif <- taxo_groups_all[!taxo_groups_all %in% c("Insecta/Lepidoptera/Rhopalocera",
                                                            "Insecta/Diptera/Syrphidae",
                                                            "Insecta/Hymenoptera/Anthophila")]
```

### Standard filters

During data cleaning there are processing steps that are applied in the same way to all taxonomic groups. These include filtering by coordinate precision, applying CoordinateCleaner tests and sampling points on a grid. We can combine all these steps into a single function that is applied to every species in each taxonomic groups. 

The filtering tables returned by the function can be combined to a single dataframe to rasterize and create maps of sampling effort and species richness after filtering. The final rasters will be the inputs of SDMs and can also be used to derive maps of species richness after data cleaning. Finally, we extract a summary table of the number of records removed at each steps which will help us validate the results of the function.

Tables of sampling effort and species richness are useful because they will help us identify the presence of sampling biases in space.

After gridding, we only have one observation per species by raster cell, so we can create species richness maps and see if the spatial pattern of bias has changed. Because there can be many species (and thus rasters) for a given taxonomic group, we will add the layers through a loop

```{r}
# Loop by taxonomic group
for (i in 1:length(taxo_groups_gbif)){
  
  # Get files
  group_i <- str_split_1(taxo_groups_gbif[i], "/")[length(str_split_1(taxo_groups_gbif[i], "/"))]
  dir.create(paste0(path_save, group_i))
  dir.create(paste0(path_save, group_i, "/Species_rasters"))
  tables_all <- list.files(paste0(path_species_data, taxo_groups_gbif[i]), 
                        pattern = ".csv", recursive = TRUE, full.names = TRUE)

  cat(paste("Starting with:", group_i, "\n"))
  cat("Applying standard filters")
  
  # Apply standard filters  
  # Save outputs
  tables_prec <- list()
  cleaning_report <- list()

  # Progress
  prbar <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed: :elapsedfull || Remaining: :eta]",
                         total = length(tables_all),
                         complete = "=", incomplete = "-", current = ">", clear = FALSE, width = 100)     


  for (species in 1:length(tables_all)){
  
    prbar$tick()
  
    # Read
    temp <- read.csv(tables_all[species], header = TRUE, encoding = "utf-8")
    
    # Remove gridded datasets from WUR provider
    temp <- temp[which(temp$institutionCode != "WUR-Alterra" |
                         is.na(temp$institutionCode)), ]
    
    # Apply standard filters
    # Filter for precision, coordinates and grid
    filter_result <- suppressWarnings(standard_filters(temp, template_raster = NaturaConnect_grid1k,
                                      precision = 1000, clean_coords = TRUE, clean_dataset = FALSE, on_grid = TRUE,
                                      outputs = c("summary", "table_prec", "final_rasters"),
                                      
                                      # CoordinateCleaner function arguments
                                      species = "species", seas_ref = EUROSTAT_Countries_1k,
                                      tests = c("equal","gbif", "institutions", "zeros", "seas")))
    
    # Retrieve outputs
    species_i <- sub(".csv", ".tif",
                     str_split_1(tables_all[species],"/")[length(str_split_1(tables_all[species], "/"))])
    tables_prec[[species]] <- filter_result[[1]]
    cleaning_report[[species]] <- filter_result[[3]]
    writeRaster(filter_result[[2]], paste0(path_save, group_i, "/Species_rasters/", species_i),
                overwrite = TRUE)
   
  }  
  
  cat("Now processing output files...\n")
  
  # Create summary dataframe
  cleaning_report_df <- map_df(cleaning_report, ~as.data.frame(t(.)))
  colnames(cleaning_report_df) <- c("Start", "Precision", "CC_coords", "CC_data", "Grid", "End")
  cleaning_report_df$File <- tables_all

  # Save
  write.csv(cleaning_report_df, paste0(path_save, group_i, "/", group_i, "_CleaningSummary.csv"), 
            row.names = FALSE)
  
  # Create maps of sampling effort and species richness in Europe

  # Combine all filtering tables to a single data.frame
  all_pfiltering <- do.call("rbind", tables_prec)

  # Convert to spatial format and reproject to NaturaConnect grid
  all_pfiltering_sp <- vect(all_pfiltering, 
                        geom = c("decimalLongitude", "decimalLatitude"), 
                        crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs", keepgeom = FALSE)
    
  all_pfiltering_sp <- terra::project(all_pfiltering_sp, NaturaConnect_grid_r)
  
  # Sampling effort: count the number of observations by cell
  sampling_effort_filt <- rasterize(all_pfiltering_sp, NaturaConnect_grid_r, field = "species", fun = length, background = NA)
  
  # Species richness: count the number of unique species by cell
  species_richness_filt <- rasterize(all_pfiltering_sp, NaturaConnect_grid_r, field = "species", fun = spec_richness, background = NA)
  
  # Save
  write.csv(all_pfiltering, paste0(path_save, group_i, "/", group_i, "_GroupObs_AfterFilt.csv"),
            row.names = FALSE)
  writeRaster(sampling_effort_filt, paste0(path_save, group_i, "/", group_i, "_SamplingEffort_AfterFilt.tif"))
  writeRaster(species_richness_filt, paste0(path_save, group_i, "/", group_i, "_SpeciesRichness_AfterFilt.tif"))
  
}
```

### Observation.org data cleaning

Ideally we would add our observation.org data to our GBIF data and clean once. However, because a few observations in the observations.org dataset are flagged as do not display at high resolution, we need to be able to differentiate the rasters produced from the two. Observation.org rasters will be saved separately and added to the GBIF rasters before the modelling part

```{r}
# Define groups for which we have data
obs_taxo_groups <- taxo_groups_gbif[-which(taxo_groups_gbif %in% c("Collembola", "Clitellata/Crassiclitellata",
                                                                   "Insecta/Orthoptera"))]

# Load observation.org filtered data
obs_org_filtered <- fread("D:/Observations_ORG/Filtered_Dataset_P0/ObservationOrg_Filtered_Data.csv")

# Loop through taxonomic groups
for (i in 1:length(obs_taxo_groups)){
  
  # Get species files
  group_i <- str_split_1(obs_taxo_groups[i], "/")[length(str_split_1(obs_taxo_groups[i], "/"))]
  dir.create(paste0(path_save_obs, group_i))
  dir.create(paste0(path_save_obs, group_i, "/Species_rasters"))
  tables_all <- list.files(paste0(path_species_data, obs_taxo_groups[i]), 
                        pattern = ".csv", recursive = TRUE, full.names = TRUE)

  cat(paste("Starting with:", group_i, "\n"))
  cat("Applying standard filters")
  
  # Apply standard filters  
  # Save outputs
  k <- 1
  cleaning_report <- list()
  tables_prec <- list()
  file_name <- c()
  obs_start <- rep(0, length(tables_all))
  is_obscure_at <- rep(0, length(tables_all))

  # Progress
  prbar <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed: :elapsedfull || Remaining: :eta]",
                         total = length(tables_all),
                         complete = "=", incomplete = "-", current = ">", clear = FALSE, width = 100)     


  for (species in 1:length(tables_all)){
  
    prbar$tick()
  
    # Read
    temp <- read.csv(tables_all[species], header = TRUE, encoding = "utf-8")
    
    # Keep observation.org records with a precision > 1 km in GBIF (others retained previously)
    temp <- temp[temp$datasetKey == "8a863029-f435-446a-821e-275f4f641165" & temp$coordinateUncertaintyInMeters > 1000, ]

    # Skip if no records
    if (!check_rows(temp)){next}
    obs_start[species] <- nrow(temp)
    
    # Match record IDs and retrieve high resolution coordinates and new precision values
    obstemp <- obs_org_filtered[link %in% temp$occurrenceID, c("naam_lat", "latitude", 
                                                               "longitude", "precision", 
                                                               "link", "obscure")]
    is_obscure_at[species] <- obstemp$obscure[1]
    temp <- merge(temp, obstemp, by.x = "occurrenceID", by.y = "link")
    if (!check_rows(temp)){next}
    
    # Apply standard filters
    # Filter for precision, coordinates and grid
    file_name[k] <- tables_all[species]
    filter_result <- suppressWarnings(standard_filters(temp, template_raster = NaturaConnect_grid1k,
                                      precision = 1000, precision_col = "precision",
                                      lon = "longitude", lat = "latitude",
                                      clean_coords = TRUE, clean_dataset = FALSE, on_grid = TRUE,
                                      outputs = c("summary", "table_prec", "final_rasters"),
                                      
                                      # CoordinateCleaner function arguments
                                      species = "species", seas_ref = EUROSTAT_Countries_1k,
                                      tests = c("equal","gbif", "institutions", "zeros", "seas")))
    
    # Retrieve outputs
    species_i <- sub("\\.csv", "_OBSORG.tif", 
                     str_split_1(tables_all[species], "/")[length(str_split_1(tables_all[species], "/"))])
    tables_prec[[k]] <- filter_result[[1]]
    cleaning_report[[k]] <- filter_result[[3]]
    writeRaster(filter_result[[2]], paste0(path_save_obs, group_i, "/Species_rasters/", species_i),
                overwrite = TRUE)
    
    k <- k + 1
   
  }  
  
  cat("Now processing output files...\n")
  
  # Create summary dataframe
  cleaning_report_df <- map_df(cleaning_report, ~as.data.frame(t(.)))
  colnames(cleaning_report_df) <- c("Obs_matched", "Precision", "CC_coords", "CC_data", "Grid", "End")
  cleaning_report_df$file_name <- file_name
  obs_report <- data.frame(Start = obs_start, obscure = is_obscure_at, name = tables_all)
  cleaning_report_df <- merge(obs_report, cleaning_report_df, 
                              by.x = "name", by.y= "file_name", all.x = TRUE)
  write.csv(cleaning_report_df, 
            paste0(path_save_obs, group_i, "/", group_i, "_CleaningSummary_OBSORG.csv"), row.names = FALSE)
  
  # Save cleaned observations and sampling effort
  all_pfiltering <- do.call("rbind", tables_prec)
  all_pfiltering_sp <- vect(all_pfiltering, 
                        geom = c("longitude", "latitude"), 
                        crs = "+proj=longlat +datum=WGS84 +no_defs +type=crs", keepgeom = FALSE)
  all_pfiltering_sp <- terra::project(all_pfiltering_sp, NaturaConnect_grid_r)
  sampling_effort_filt <- rasterize(all_pfiltering_sp, NaturaConnect_grid_r, 
                                    field = "species", fun = length, background = NA)
  species_richness_filt <- rasterize(all_pfiltering_sp, NaturaConnect_grid_r, 
                                     field = "species", fun = spec_richness, background = NA)
  write.csv(all_pfiltering, paste0(path_save_obs, group_i, "/", group_i, "_GroupObs_AfterFilt_OBSORG.csv"),
            row.names = FALSE)
  writeRaster(sampling_effort_filt, 
              paste0(path_save_obs, group_i, "/", group_i, "_SamplingEffort_AfterFilt_OBSORG.tif"))
  writeRaster(species_richness_filt, 
              paste0(path_save_obs, group_i, "/", group_i, "_SpeciesRichness_AfterFilt_OBSORG.tif"))
  
  }
```

## Adding information on the species' origin

Because GBIF data are extracted using a range including parts if north Africa and west Asia, some of the species in the GBIF downloads may not be European species. We can find the number of observations in the EU+ study area to calculate the percentage of non-EU observations using the raster layers out putted for each species.

For our EU+ extent we can use Yue's land-use map. then we go through our species layers and count the number of pixels retained after masking.

We can also assess the number of observations by species located in urban areas or freshwater covered areas that do not have soil data available.

Finally, we can find the number of observations of each species in Ireland and England as we will use these areas to create a fake evaluation set for selecting our PA set.

```{r}
# Load land-use layer
EUp_zones <- rast(paste0(path_GIS_data, "LandUse/LandCover_Yue_WithConiferousForest.tif"))

# Convert 0s to NAs
EUp_zones[EUp_zones <= 0] <- NA

# Extend to NaturaConncet extent
EUp_zones <- terra::extend(EUp_zones, NaturaConnect_grid_r)

# Mask grid raster
EUp_zones <- mask(rast(NaturaConnect_grid1k), EUp_zones)
EUp_zones[!is.na(EUp_zones)] <- 1
```

```{r}
# Load  soil layer
soil_data <- rast(paste0(path_GIS_data, "Prepped_layers/VarPart_V/Soil_Variables_All.tif"))[["phh2o"]]
soil_data[soil_data > 0] <- 1
soil_data[soil_data <= 0] <- NA
```
```{r}
# Load Ireland and UK polygon (extracted from EUROSTAT)
IE_UK_poly <- project(vect(paste0(path_GIS_data, 
                                  "CountryBorders_EUROSTAT/EUROSTAT_IrelandUK_buffered1k_WGS84.shp")), NaturaConnect_grid_r)
IE_UK_rast <- rasterize(IE_UK_poly, NaturaConnect_grid_r, field = "FID", background = NA)
```


Get final rasters and find number of final observations by species, observations in the EU and observations with available soil data

```{r}
for (i in 1:length(taxo_groups_all)){
  
  # Get species
  group_i <- str_split_1(taxo_groups_all[i], "/")[length(str_split_1(taxo_groups_all[i], "/"))]
  group_species <- list.files(paste0(path_save, group_i, "/Species_rasters/"))
  cat(paste0("Starting with: ", group_i, "\n"))
  
  # Prepare summary data frame
  cleaned_obs <- rep(0, length(group_species))
  EUp_obs <- rep(0, length(group_species))
  soil_obs <- rep(0, length(group_species))
  IEUK_obs <- rep(0, length(group_species))
  
  # Progress bar
  prbar <- progress_bar$new(format = "(:spin) [:bar] :percent [Elapsed: :elapsedfull || Remaining: :eta]",
                            total = length(group_species),
                            complete = "=", incomplete = "-", current = ">", clear = FALSE, width = 100)
  
  # Create directories
  dir.create(paste0(path_final, group_i, "/Species_rasters"), recursive = TRUE)
  
  # Combine all species rasters
  for (species in 1:length(group_species)){
    
    prbar$tick()
    species_i <- sub("\\.tif", "", group_species[species])
    
    # Read species raster
    species_raster <- combine_rasters(paste0(group_i, "/Species_rasters/", species_i),
                                      NaturaConnect_grid_r, fun = "max") 
    # Number of final observations 
    cleaned_obs[species] <- global(species_raster, fun = "sum", na.rm = TRUE)[1, 1]
    
    # Mask with EUp
    species_raster_EU <- mask(species_raster, EUp_zones, maskvalues = NA)
    
    # Number of EU observations
    EUp_obs[species] <- global(species_raster_EU, fun = "sum", na.rm = TRUE)[1, 1] 
    
    # Mask with available soil data
    species_raster_soil <- mask(species_raster, soil_data, maskvalues = NA)
    
    # Number of observations with soil data available
    soil_obs[species] <- global(species_raster_soil, fun = "sum", na.rm = TRUE)[1, 1]
    
    # Mask with Ireland  UK
    species_raster_IEUK <- mask(species_raster, IE_UK_rast, maskvalues = NA)
    
    # Number of observations in Ireland & the UK
    IEUK_obs[species] <- global(species_raster_IEUK, fun = "sum", na.rm = TRUE)[1, 1]
    
    # Write final species raster
    ifelse(is.na(cleaned_obs[species]), next, 
           writeRaster(species_raster, 
                       paste0(path_final, group_i, "/Species_rasters/", species_i, ".tif")))
    
  }
  
  # Save group df results
  group_results <- data.frame(Species = sub("\\.tif", "", group_species), 
                              All = cleaned_obs, 
                              EUp = EUp_obs,
                              Soil = soil_obs,
                              IEUK = IEUK_obs)
  
  write.csv(group_results, paste0(path_results, "SummarypostEUP/", group_i, "_FilteringResults.csv"), 
            row.names = FALSE)
  
  # Get sum of sampling effort
  sampling_effort <- combine_rasters(paste0(group_i, "/", group_i, "_SamplingEffort_AfterFilt"),
                                     NaturaConnect_grid_r, fun = "sum")
  
  # Path to output
  output_path <- paste0(path_final, group_i, "/", group_i, "_SamplingEffort_All.tif")
  
  # Deal with bees and wasps
  writeRaster(sampling_effort, output_path)
  
  
}
```




```{r}
# Species lists for NaturaConnect
# Read final list of species by group
taxo_groups_new <- c("Anthophila", "Araneae", "Carabidae", "Formicidae", "Orthoptera", "Odonata", "Lepidoptera",
                     "Gastropoda", "Ichneumonoidea", "Syrphidae", "Crassiclitellata", "Collembola")

for (i in 1:length(taxo_groups_new)){
  
  # Read summary file, for bees and wasps we will need to merge the tables for all fmilies
  if (taxo_groups_new[i] == "Anthophila"){
    
    all_dfs <- list()
    families <- c("Andrenidae", "Halictidae", "Colletidae", "Apidae", "Melittidae", "Megachilidae")
    for (f in 1:length(families)){
      
      all_dfs[[f]] <- read.csv(paste0(path_results, "SummarypostEUP/", families[f], "_cleaning_end_obs.csv"),
                                 header = TRUE)
      all_dfs[[f]]$Group <- rep(families[f], nrow(all_dfs[[f]]))
      
    }
    
    tmp <- do.call("rbind", all_dfs)
    
  } else if (taxo_groups_new[i] == "Ichneumonoidea"){
    
    all_dfs <- list()
    families <- c("Braconidae", "Ichneumonidae")
    for (f in 1:length(families)){
      
      all_dfs[[f]] <- read.csv(paste0(path_results, "SummarypostEUP/", families[f], "_cleaning_end_obs.csv"),
                                 header = TRUE)
      all_dfs[[f]]$Group <- rep(families[f], nrow(all_dfs[[f]]))
      
    }
    
    tmp <- do.call("rbind", all_dfs)

    
  } else {
    
      tmp <- read.csv(paste0(path_results, "SummarypostEUP/", taxo_groups_new[i], "_cleaning_end_obs.csv"),
                                 header = TRUE)
      
      tmp$Group <- rep(taxo_groups_new[i], nrow(tmp))
    
  }
  
  tmp[is.na(tmp)] <- 0
  tmp <- tmp[tmp$All >= 20, c("Species", "Group")]
  
  write.csv(tmp, paste0(path_results, "SpeciesLists/27-06-23/", taxo_groups_new[i], "_species_list.csv"), row.names = FALSE)
  
}

# All species for selected groups
sel_taxa <- c("Araneae", "Carabidae", "Formicidae", "Orthoptera", "Odonata", 
              "Lepidoptera", "Gastropoda", "Syrphidae")

all_species <- list()

for (i in 1:length(sel_taxa)){
  
  all_species[[i]] <- read.csv(paste0(path_results, "SpeciesLists/27-06-23/", sel_taxa[i], "_species_list.csv"), header = TRUE)
  
}

all_species <- do.call("rbind", all_species)
write.csv(all_species, paste0(path_results, "SpeciesLists/27-06-23/Provisional_invertebrate_species_list.csv"), row.names = FALSE)
```

